Model Search in Relational Domains

### 2016-10-24 domains.tex
final proof-read

%-------------------------------------------------------------------------------
\subsection{Program Representation}  
\Omit {
\begin{wrapfigure}{r}{3cm}
\[
\begin{array}{rcl}
g_0 &=& \true \\
\multicolumn{3}{l}{0 \leq v\leq N} \\
g_1 &=& g_0\wedge c\\
x_0 &=& v \\
x_1 &=& -v \\
x_2 &=& g_1?x_0:x_1 \\
g_2 &=& g_1 \vee g_0\wedge \neg c\\
z &=& x_2 * x_2 \\
\multicolumn{3}{l}{z<0 \wedge g_2}
\end{array}
\]
\caption{\label{fig:ssa}
SSA for the example in Figure~\ref{fig:example}
}
\end{wrapfigure}
}
\begin{figure}[t]
\center
\scriptsize
\begin{tabular}{l|l}
\hline
 SSA & Abstract Conflict Graph \\
\hline
\begin{minipage}{4.50cm}
$\begin{array}{rcl}
\constraints:= (g_0 = (0 \leq v\leq N)) \\
(g_1 = (g_0 \wedge c)) \wedge (x_0 = v) \wedge \\
(x_1 = -v) \wedge (x_2 = g_1?x_0:x_1) \wedge \\
(g_2 = (g_1 \vee g_0\wedge \neg c)) \wedge \\
(z  = x_2 * x_2) \wedge (g_2 \wedge z<0)
\end{array}$
\end{minipage}
&
\begin{minipage}{5.2cm}
\centering
\scalebox{.45}{\import{figures/}{learning.pspdftex}}
\end{minipage}
\\
\hline
\end{tabular}
\caption{SSA and abstract conflict graph}
\label{ssa}
\end{figure}
%
In this paper, we consider \emph{bounded programs} with safety
properties given as assertions $\assertions$ in the program.
%
A bounded program is obtained by a program transformation that unfolds
loops and recursions a finite number of times. The result of this
transformation is represented as set
$\constraints=\program\cup\{\neg \bigwedge_{\assertion\in\assertions} \assertion\}$
where $\program$ are the statements of the transformed program in the
form of constraints as typically obtained by translating the program
into single static assignment (SSA) form via a data flow analysis.
%
Figure~\ref{ssa} gives the SSA constraints for the program in
Figure~\ref{fig:example}.
%
Assignments such as \texttt{x=v;} become equalities $x_1=v$ where the
left-hand side variable gets a fresh name (subscript).
%
Control flow is encoded into guard variables such as $g_1=g_0\wedge c$.
%
Data flow joins become conditional expressions, e.g.\ $x_3=g_1?x_1:x_2$.
%
The assertions $\assertions$ are constraints of the form $g_3
\Rightarrow z\geq 0$, for instance, meaning that if $g_3$ holds
(i.e.\ the assertion is reachable) then the assertion must hold.
%
We~denote $\vars$ the set of variables occurring in $\constraints$.  
Based on the above program representation, we define a safety equation as, 
$\formula:= \bigwedge_{\constraint\in\constraints} \constraint$. The formula $\formula$ is 
unsatisfiable if and only if the program is safe.
%
%
%For example, figure~\ref{swssa} presents the program analysis constraints for 
%a C program in Static Single Assignment (SSA) form.

@@

%-------------------------------------------------------------------------------
\subsection{Abstract Domain}
%-------------------------------------------------------------------------------
SAT solvers operate on partial assignment domains~\cite{sas12,dhk2013-popl}.  
We present an instantiation of ACDCL as program analyser that uses relational
domains.  
% 
%In our prototype implementation and for illustrative purposes 
In this paper we use a product domain $\domain[\vars]=\booldomain^{|\boolvars|}\times\reldomain[\numvars]$ where
$\booldomain$ is the Boolean domain
$\langle\{\true,\false,\bot,\top\},\Rightarrow,\wedge,\vee\rangle$, 
$\boolvars$ are the Boolean variables in the program, 
and $\reldomain$ is a \emph{relational} domain over the 
numerical (bitvector) variables $\numvars$ of their respective
data types (currently, signed and unsigned integers are supported).
%
We instantiate $\reldomain$ with the template polyhedra domain.

For notational convenience we will denote elements of
$\booldomain^{|\boolvars|}$ by their concretisations to propositional formulae,
For example, the program in Figure~\ref{ssa} has four Boolean variables
written as the vector $(g_0,g_1,c,g_2)$; then we will denote an abstract value $(\top, \false, \true, \top) \in \booldomain^4$ as $\neg g_1 \wedge c$.

%-------------------------------------------------------------------------------
\paragraph{\textbf{Template Polyhedra Abstract Domain}}
%
An abstract value of the template polyhedra domain~\cite{sriram}
represents a set $\numconcval$ of values of the vector $\vec{\numvar}$ of numerical variables $\numvars$. For
example, in the program in Figure~\ref{ssa}, we have four numerical
variables, written as the vector $\vec{\numvar}:=(x_0,x_1,x_2,z)$.  We can
represent sets of values for $\vec{\numvar}$ such that
$\mat{C}\vec{\numvar}\leq\vec{\numabsval}$ for a fixed coefficient matrix $\mat{C}$
and a constant vector $\vec{\numabsval}$. The domain of $\vec{\numabsval}$ is augmented
by a special element $\bot$ to denote the minimal element of the
lattice.
%
The abstraction function is defined as $\alpha(\numconcval) = \min \{\vec{\numabsval}\mid
\mat{C}\vec{\numvar}\leq\vec{\numabsval}, \vec{\numvar}\in \numconcval\}$ where $\min$ is applied
component-wise. There are several optimisation-based
techniques~\cite{sriram,GS07b,BJKS15} for computing the domain
operations, such as meet ($\meet$), join ($\sqcup$), etc. In our implementation we
use the strategy iteration approach available in 2LS~\cite{BJKS15}.
%
The concretisation $\gamma(\vec{\numabsval})$ is the set $\{\vec{\numvar}\mid
\mat{C}\vec{\numvar}\leq\vec{\numabsval}\}$; $\gamma(\bot)=\emptyset$, i.e.\ the empty polyhedron.
%
For notational convenience we will use conjunctions of linear
inequalities, e.g.\ $x_1\geq 0 \wedge x_1-z\leq 30$ to denote the
abstract value $\vec{\numabsval}=\vecv{0}{30}$ of the template polyhedra domain
with $\mat{C}=\qmat{-1}{0}{1}{-1}$ and $\vec{\numvar}=\vecv{x_1}{z}$;
$\true$ is $\top$ and $\false$ is $\bot$.  Template polyhedra can
express other domains, such as interval and octagon domains.

@@

%-------------------------------------------------------------------------------
\subsection{Properties of Abstract Domains}
%-------------------------------------------------------------------------------
%
An important property of a clause learning SAT solver is that each
non-singleton element of the partial assignment domain can be 
decomposed into a set of precisely complementable singleton elements.  
This property of the partial assignment domain is required to learn 
elements that help to guide the model search away from the
conflicting region of the search space.  The complementation operator 
in abstract domains is different from the notion of precise complements. 
%
Most numerical abstract domains, such as intervals and octagons lack
complements, but they can be represented as intersections of
half-spaces, each of which has a complement. 
%
% We identify specific properties of domain elements necessary for 
% abstract model search and learning in abstractions.  
We formalise this in the sequel.
%
\begin{definition} 
A \emph{meet irreducible} $m$ in a complete lattice 
structure $A$ is an element with the following property.
\begin{equation}
\forall m_1, m_2 \in A: m_1 \meet m_2 = m \implies (m = m_1 \lor m = m_2), m \neq \top  
\end{equation}
\end{definition}
%
%Meet irreducibles in ACDCL correspond to the concept of literals in a SAT solver.
The meet irreducibles in the Boolean domain $\booldomain$ for a variable $x$ are 
$x$ and $\neg x$. The meet 
irreducibles in the template polyhedra domain are all elements 
that concretise to half-spaces, i.e., they can be represented 
by a single inequality. For the interval domain, these are 
\ $x \leq d$ or $x \geq d$ for constants $d$.

>>>>

%-------------------------------------------------------------------------------
\subsection{Program Representation}  
\Omit {
\begin{wrapfigure}{r}{3cm}
\[
\begin{array}{rcl}
g_0 &=& \true \\
\multicolumn{3}{l}{0 \leq v\leq N} \\
g_1 &=& g_0\wedge c\\
x_0 &=& v \\
x_1 &=& -v \\
x_2 &=& g_1?x_0:x_1 \\
g_2 &=& g_1 \vee g_0\wedge \neg c\\
z &=& x_2 * x_2 \\
\multicolumn{3}{l}{z<0 \wedge g_2}
\end{array}
\]
\caption{\label{fig:ssa}
SSA for the example in Figure~\ref{fig:example}
}
\end{wrapfigure}
}
\begin{figure}[t]
\center
\scriptsize
\begin{tabular}{l|l}
\hline
 SSA & Abstract Conflict Graph \\
\hline
\begin{minipage}{4.50cm}
$\begin{array}{rcl}
\constraints:= (g_0 = (0 \leq v\leq N)) \\
(g_1 = (g_0 \wedge c)) \wedge (x_0 = v) \wedge \\
(x_1 = -v) \wedge (x_2 = g_1?x_0:x_1) \wedge \\
(g_2 = (g_1 \vee g_0\wedge \neg c)) \wedge \\
(z  = x_2 * x_2) \wedge (g_2 \wedge z<0)
\end{array}$
\end{minipage}
&
\begin{minipage}{5.2cm}
\centering
\scalebox{.45}{\import{figures/}{learning.pspdftex}}
\end{minipage}
\\
\hline
\end{tabular}
\caption{SSA and abstract conflict graph}
\label{ssa}
\end{figure}
%
We consider \emph{bounded programs} with safety
properties given as assertions $\assertions$ in the program.
%
A bounded program is obtained by a program transformation that unfolds
loops and recursions a finite number of times. The result of this
transformation is represented by a set
$\constraints=\program\cup\{\neg \bigwedge_{\assertion\in\assertions} \assertion\}$
where $\program$ contains the statements of the transformed program in the
form of constraints as typically obtained by translating the program
into single static assignment (SSA) form via a data flow analysis.
%
Figure~\ref{ssa} gives the SSA constraints for the program in
Figure~\ref{fig:example}.
%
Assignments such as \texttt{x=v;} become equalities $x_1=v$ where the
left-hand side variable gets a fresh name (subscript).
%
Control flow is encoded into guard variables such as $g_1=g_0\wedge c$.
%
Data flow joins become conditional expressions, e.g.\ $x_3=g_1?x_1:x_2$.
%
The assertions $\assertions$ are constraints of the form $g_3
\Rightarrow z\geq 0$, for instance, meaning that if $g_3$ holds
(i.e.\ the assertion is reachable) then the assertion must hold.
%
We~denote $\vars$ the set of variables occurring in $\constraints$.  
Based on the above program representation, we define a safety equation as, 
$\formula:= \bigwedge_{\constraint\in\constraints} \constraint$. The formula $\formula$ is 
unsatisfiable if and only if the program is safe.
%
%
%For example, figure~\ref{swssa} presents the program analysis constraints for 
%a C program in Static Single Assignment (SSA) form.

@@

%-------------------------------------------------------------------------------
\subsection{Abstract Domain}
%-------------------------------------------------------------------------------
SAT solvers operate on partial assignment domains~\cite{sas12,dhk2013-popl}.  
We present an instantiation of ACDCL as program analyser that uses relational
domains.  
% 
%In our prototype implementation and for illustrative purposes 
In this paper we use a product domain $\domain[\vars]=\booldomain^{|\boolvars|}\times\reldomain[\numvars]$ where
$\booldomain$ is the Boolean domain
$\langle\{\true,\false,\bot,\top\},\Rightarrow,\wedge,\vee\rangle$, 
$\boolvars$ are the Boolean variables in the program, 
and $\reldomain$ is a \emph{relational} domain over the 
numerical (bitvector) variables $\numvars$ of their respective
data types (currently, signed and unsigned integers are supported).
%
We instantiate $\reldomain$ with the template polyhedra domain~\cite{sriram}.

For notational convenience we will denote elements of
$\booldomain^{|\boolvars|}$ by their concretisations to propositional formulae.  
For example, the program in Figure~\ref{ssa} has four Boolean variables
written as the vector $(g_0,g_1,c,g_2)$. Thus, we will denote an abstract 
value $(\top, \false, \true, \top) \in \booldomain^4$ as $\neg g_1 \wedge c$.

%-------------------------------------------------------------------------------
\paragraph{\textbf{Template Polyhedra Abstract Domain}}
%
An abstract value of the template polyhedra domain~\cite{sriram}
represents a set $\numconcval$ of values of the vector $\vec{\numvar}$ of numerical variables $\numvars$. For
example, in the program in Figure~\ref{ssa}, we have four numerical
variables, written as the vector $\vec{\numvar}:=(x_0,x_1,x_2,z)$.  We can
represent sets of values for $\vec{\numvar}$ such that
$\mat{C}\vec{\numvar}\leq\vec{\numabsval}$ for a fixed coefficient matrix $\mat{C}$
and a constant vector $\vec{\numabsval}$. The domain of $\vec{\numabsval}$ is augmented
by a special element $\bot$ to denote the minimal element of the
lattice.
%
The abstraction function is defined as $\alpha(\numconcval) = \min \{\vec{\numabsval}\mid
\mat{C}\vec{\numvar}\leq\vec{\numabsval}, \vec{\numvar}\in \numconcval\}$ where $\min$ is applied
component-wise. There are several optimisation-based
techniques~\cite{sriram,GS07b,BJKS15} for computing the domain
operations, such as meet ($\meet$), join ($\sqcup$), etc in the template
polyhedra domain. In our implementation, we use the strategy 
iteration approach available in 2LS~\cite{BJKS15}.
%
The concretisation $\gamma(\vec{\numabsval})$ is the set $\{\vec{\numvar}\mid
\mat{C}\vec{\numvar}\leq\vec{\numabsval}\}$; $\gamma(\bot)=\emptyset$, i.e.\ the empty polyhedron.
%
For notational convenience we will use conjunctions of linear
inequalities, e.g.\ $x_1\geq 0 \wedge x_1-z\leq 30$ to denote the
abstract value $\vec{\numabsval}=\vecv{0}{30}$ of the template polyhedra domain
with $\mat{C}=\qmat{-1}{0}{1}{-1}$ and $\vec{\numvar}=\vecv{x_1}{z}$;
$\true$ is $\top$ and $\false$ is $\bot$.  Template polyhedra can
express other domains, such as interval and octagon domains.

@@

%-------------------------------------------------------------------------------
\subsection{Properties of Abstract Domains}
%-------------------------------------------------------------------------------
%
An important property of a clause learning SAT solver is that each
non-singleton element of the partial assignment domain can be 
decomposed into a set of precisely complementable singleton
elements~\cite{dhk2013-popl}.  This property of the partial 
assignment domain is required to learn 
elements that help to guide the model search away from the
conflicting region of the search space.  The complementation operator 
in abstract domains is different from the notion of precise complements. 
%
Most numerical abstract domains, such as intervals and octagons lack
complements, but they can be represented as intersections of
half-spaces, each of which has a complement. 
%
% We identify specific properties of domain elements necessary for 
% abstract model search and learning in abstractions.  
We formalise this in the sequel.
%
\begin{definition} 
A \emph{meet irreducible} $m$ in a complete lattice 
structure $A$ is an element with the following property.
\begin{equation}
\forall m_1, m_2 \in A: m_1 \meet m_2 = m \implies (m = m_1 \lor m = m_2), m \neq \top  
\end{equation}
\end{definition}
%
%Meet irreducibles in ACDCL correspond to the concept of literals in a SAT solver.
The meet irreducibles in the Boolean domain $\booldomain$ for a variable $x$ are 
$x$ and $\neg x$. The meet 
irreducibles in the template polyhedra domain are all elements 
that concretise to half-spaces, i.e., they can be represented 
by a single inequality. For the interval domain, these are 
\ $x \leq d$ or $x \geq d$ for constants $d$.
